---
title: "LDSR 2021 Re-analysis (Refactored)"
author: "Rowland Pettit"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Purpose
Refactor of the 2021 LDSR re-analysis to make intent, steps, and decisions explicit. This version removes local paths, adds commentary, and uses placeholders pointing to public resources.

# Background
- Remove ±500kb windows around smoking-associated SNPs to assess whether lung cancer correlations persist independently of smoking loci.
- Use LDSC to estimate h2 and rg with UKB traits; compare filtered vs. unfiltered results.

# Setup
```{r setup, message=FALSE}
library(dplyr)
library(readr)
library(stringr)
library(ggplot2)
```

# Inputs (placeholders)
```{r inputs, eval=FALSE}
# Smoking SNPs (from Liu 2019 NG; provide a CSV of significant loci)
smoking_snps <- readr::read_csv("path/to/Liu_2019_NG_smoking_snps.csv")
excluded_regions <- smoking_snps %>% transmute(CHR = Chr,
                                               Start = Pos - 500000,
                                               End   = Pos + 500000)

# TRICL/OncoArray summary stats (raw), one file per outcome
tricl_files <- list(
  overall = "path/to/tricl/overall.txt",
  adeno   = "path/to/tricl/adeno.txt",
  small   = "path/to/tricl/small.txt",
  squam   = "path/to/tricl/squam.txt",
  ever    = "path/to/tricl/ever.txt",
  never   = "path/to/tricl/never.txt"
)
```

# Helper: exclude windows
```{r exclude-helper}
exclude_windows <- function(stat, excluded_regions) {
  stopifnot(all(c("CHR", "position") %in% names(stat)))
  stopifnot(all(c("CHR", "Start", "End") %in% names(excluded_regions)))
  idx <- logical(nrow(stat))
  for (j in seq_len(nrow(excluded_regions))) {
    r <- excluded_regions[j, ]
    idx <- idx | (stat$CHR == r$CHR & stat$position >= r$Start & stat$position <= r$End)
  }
  stat[!idx, ]
}
```

# Process TRICL outcomes (filter vs. unfiltered)
```{r process-tricl, eval=FALSE}
process_tricl <- function(path, out_filtered) {
  cols <- c("chr:pos","rs_number","CHR","position","reference_allele","effect_allele",
            "EAF","OR_fixed","StdError_fixed","Pvalue_fixed","L95_fixed","U95_fixed",
            "phete(Q)_Pvalue","I2","N_study","N_Cases","N_Controls","Effects",
            "OR_random","StdError_random","Pvalue_random","L95_random","U95_random")
  x <- readr::read_tsv(path, col_names = FALSE, show_col_types = FALSE, progress = FALSE, skip = 1)
  names(x) <- cols
  if (out_filtered) x <- exclude_windows(x, excluded_regions)
  x
}

tricl <- lapply(names(tricl_files), function(k) list(
  raw = process_tricl(tricl_files[[k]], FALSE),
  fil = process_tricl(tricl_files[[k]], TRUE)
))
names(tricl) <- names(tricl_files)
```

# LDSC: Munge and run (command templates)
```{r ldsc-templates}
ldsc_dir <- "~/code/ldsc"              # where ldsc.py lives
ref_dir  <- "~/refs/ldsc/eur_w_ld_chr" # 1KG EUR LD scores

munge_cmd <- function(sumstats, outstem) sprintf(
  "python %s/munge_sumstats.py --sumstats %s --out %s --N %d --merge-alleles %s/w_hm3.snplist",
  ldsc_dir, sumstats, outstem, 100000, ref_dir
)

h2_cmd <- function(munged, outstem) sprintf(
  "python %s/ldsc.py --h2 %s --ref-ld-chr %s/ --w-ld-chr %s/ --out %s",
  ldsc_dir, munged, ref_dir, ref_dir, outstem
)

rg_cmd <- function(munged1, munged2, outstem) sprintf(
  "python %s/ldsc.py --rg %s,%s --ref-ld-chr %s/ --w-ld-chr %s/ --out %s",
  ldsc_dir, munged1, munged2, ref_dir, ref_dir, outstem
)
```

# Parse LDSC outputs (example)
```{r parse-ldsc, eval=FALSE}
# Example: read h2 results and RG logs, then create comparison tables
h2 <- readr::read_tsv("trait1_h2.results")
# rg often appears in logs; parse by regex or use helper if available
```

# Visualization templates
```{r viz-templates}
plot_rg_heatmap <- function(df, rg_col = "rg", p_col = "p", lower = -0.8, upper = 0.8) {
  df$neglog10 <- -log10(as.numeric(df[[p_col]]))
  ggplot(df, aes(x = Outcome, y = Trait, fill = !!rlang::sym(rg_col))) +
    geom_tile() +
    geom_text(aes(label = paste0(round(!!rlang::sym(rg_col), 2), ifelse(df$neglog10 > 5, " *", ""))), size = 3) +
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limits = c(lower, upper)) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    labs(x = "Outcomes", y = "Traits", fill = "rg")
}
```

# Notes
- Ensure ancestry-matched LD reference and GWAS.
- Use effective N for case-control.
- Intercepts near 1 suggest minimal confounding; large deviations warrant caution.

# References
- Bulik-Sullivan BK, et al. Nat Genet (2015)
- LDSC: https://github.com/bulik/ldsc

---

## Results interpretation checklist
- Munge logs: check SNP count retained; confirm HM3 merge applied.
- h2 intercepts: near 1 indicates modest confounding; inspect gcov intercept in rg.
- Annotations: for partitioning, ensure baselineLD matches reference panel and ancestry.
- Effective N: correct case/control sample sizes for binary traits.
- Trait definitions: harmonize naming across UKB manifests to avoid mismatches.
- Multiple testing: adjust across many traits (Bonferroni or FDR).
- Sensitivity: replicate with/without exclusion of smoking-associated windows to assess mediation.
